---
title: "STAT 345 Final Project - Thanos Snap"
author: "STAT 345"
output: html_document
---

Expected Submission: You will generate a well-written R Markdown report that addresses the following prompts. This R Markdown report should source your analysis code and only display top-level, abstracted code _if/when appropriate_. Choices made during the analysis project should be described and justified in the report. 


Advice for getting started:

- Start a conversation chain with your group members. Email is a great place to start, but other options exist (texting, social media platforms, etc.). Set some reasonable expectations for how and when you plan to respond to conversations. It is likely going to be unacceptable to only respond once per week, for example, but also likely unacceptable to expect a response within the hour. Have an honest conversation with your group members about this issue.
- Start the project from a "top-down design" perspective. So, determine what the major steps of the process are, and determine what the inputs and outputs are for each of those steps (the output of step 1 will likely be the input for step 2, for example). This step is much like writing an outline for a paper before you start to write the paper itself, only much more valuable in a group setting. 
- Once you have a sense of the big picture (the outline), determine some "due dates" for each step of the process. Work backwards from the final submission date, giving yourselves time to work on each of the parts as needed. Given the top-down design, you should be able to "divide and conquer" -- working on parts of the project that depend on earlier steps.
- Decide how you plan to share code with each other. Using Git and GitHub is a really good choice here. If not, perhaps some form of shared directory online. In a worst-case scenario, email could also work. 
- Be prepared to give Dr. Baumann (at least) weekly updates on your progress. Some of this can be done via email, but discuss with your group about times you are each available to meet online as well (in an office-hour-like setting). Feel free to request meetings with Dr. Baumann to get help.

**General advice:** Get started early. If you wait to the last minute, it will not go well. For this project, you may find yourself spending a reasonable amount of time _searching_ for help.

1. _Genetic Diversity_ Familiarize yourself with some concepts regarding genetic diversity. There are many resources online for this, such as [https://www.ck12.org/c/biology/genetic-drift/lesson/Population-Size-and-Genetic-Drift-Advanced-BIO-ADV/#:~:text=Genetic%20drift%20can%20result%20in,effect%20and%20the%20founder%20effect.](https://www.ck12.org/c/biology/genetic-drift/lesson/Population-Size-and-Genetic-Drift-Advanced-BIO-ADV/#:~:text=Genetic%20drift%20can%20result%20in,effect%20and%20the%20founder%20effect.). For simplicity, you only need to consider one allele for this project. Common concepts include alleles, allele frequency, genetic drift, generations, population size, and bottleneck events.

2. _Genetic Drift_ Simulate random genetic drift. Your simulation should be able to accommodate population size, starting allele frequencies, and other relevant factors found in part 1. You should also consider running these simulations many times and summarizing the results. Create a complete, polished visualization for your simulation results. Summary statistics describing the simulation should also be developed.

3. _User Friendly_ Create an interactive document using Shiny that allows for a user to specify variable values and display the visualization based on part 2. Your document should also display other relevant summary statistics either in graphic form or tabular form. Additional user-friendly features should be considered. The websites below include basic simulations for genetic drift and bottleneck events. Feel free to use these as inspiration. 

https://www.biologysimulations.com/genetic-drift-bottleneck-event
http://www.biology.arizona.edu/evolution/act/drift/about.html
https://virtualbiologylab.org/population-genetics/



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installed Packages
```{r}
library(tidyverse)
library(shiny)
```

## Bottleneck Function
```{r}
# This function takes a population and randomly removes a number of observations from it.
# The number of observations removed will be no less than 10% and no more than 90%.
# Inputs: population = a vector representing individuals in the population
# Outputs: a data frame with the surviving population and the percent that was removed
remove_random_pct <- function(population) {
  n <- length(population) # Gets the size of the population

  pct <- runif(1, min = 0.1, max = 0.9) # Randomly generates a number between 0.1 and 0.9

  num_remove <- floor(n * pct) # Calculates the number of observations to remove
  remove_indices <- sample(seq_len(n), size = num_remove) # Selects random indices to remove

  result <- population[-remove_indices] # Removes the observations considered "dead"

  final_result <- data.frame(result = result, pct_removed = pct) # Creates a data frame of survivors and percent removed

  return(final_result)
}
```

## Genetic Drift Simulation Function
```{r}
# This function simulates genetic drift over multiple generations, which will show how allele frequencies change over time.
# The inputs are an integer for population size, a vector of frequencies, and an integer for the number of generations.
# The output is a data frame with generations, alleles, and the frequencies of the alleles.
drift_simulation <- function(new_population_size, frequencies, generations) {
  
  ##Create generation 1 using the starting allele frequencies
  alleles <- rep(names(frequencies), times = round(frequencies * new_population_size))
  
  ##Fix rounding issues so the population size stays consistent
  diff <- new_population_size - length(alleles)
  if (diff > 0) {
    alleles <- c(alleles, sample(names(frequencies), diff, replace = TRUE, prob = frequencies))
  } else if (diff < 0) {
    alleles <- sample(alleles, new_population_size)
  }

  ##Shuffle so alleles are randomly distributed
  population <- sample(alleles)

  ##Store generation 1 results
  results <- data.frame(generation = 1, allele =names(frequencies), frequency= as.numeric(table(factor(population, levels = names(frequencies))))/new_population_size)

  ##Run the simulation for the remaining generations
  for (gen in 2:generations) {
    ##Randomly sample the population to simulate genetic drift
    population <- sample(population,size = new_population_size, replace = TRUE)
    ##Calculate new allele frequencies
    freq <- as.numeric(table(factor(population, levels =names(frequencies))))/new_population_size
    ##Add results to the data frame
    results <- rbind(results, data.frame(generation = gen, allele = names(frequencies), frequency = freq))
  }

  return(results)
}
```

## Run Multiple Simulations Function
```{r}
# This function runs the genetic drift simulation multiple times and combines results.
# Inputs: num_simulations = integer for how many times to run the simulation,
#         population_size = integer for post-bottleneck population size,
#         frequencies = named vector of starting allele frequencies,
#         generations = integer for number of generations to simulate
# Outputs: a combined data frame from all simulations with an added simulation column
run_simulations <- function(num_simulations, population_size, frequencies, generations) {
  # map_dfr runs the simulation repeatedly and combines results into one data frame
  # The anonymous function(sim) tracks which simulation number each row came from
  results <- map_dfr(1:num_simulations, function(sim) {
    drift_simulation(population_size, frequencies, generations) %>%
      mutate(simulation = sim)
  })

  return(results)
}
```

## Summary Statistics Function
```{r}
# This function calculates summary statistics for allele frequencies at the final generation across all simulations.
# Inputs: results data frame with generation, allele, frequency, and simulation columns
# Outputs: a summarized data frame showing mean, standard deviation, fixation rate, and loss rate of final generation.
summary_stats <- function(results) {
  results %>%
    ##Filter for the final generation
    filter(generation == max(generation)) %>%
    ##Group by allele
    group_by(allele) %>%
    ##Calculate summary statistics
    summarise(
      Mean_Final_Frequency = round(mean(frequency), 4), #The average ending frequency across all simulations
      SD_Final_Frequency = round(sd(frequency), 4), #how much the ending frequency varied between simulations
      Fixation_Rate = round(mean(frequency == 1.0), 4), #the proportion of simulations where an allele reached a frequency  of 1.0 (meaning it took over completely)
       Loss_Rate = round(mean(frequency == 0.0), 4) #the proportion of simulations where an allele hit 0.0 frequency (meaning it died out/went extinct)
    )
}
```

## Shiny App
```{r}
# UI
ui <- fluidPage(
  titlePanel("Genetic Drift Simulation"),

  ##Create side bar and layout for users to pick what they want.
  sidebarLayout(
    sidebarPanel(
      numericInput("pop_size", "Population Size:", value = 1000, min = 10),
      ##This is haploid genetic drift.
      sliderInput("brown", "Brown Frequency:", min = 0, max = 1, value = 0.5, step = 0.01),
      sliderInput("blue", "Blue Frequency:", min = 0, max = 1, value = 0.3, step = 0.01),
      helpText("Green frequency = 1 - Brown - Blue. Must sum to 1 or less."),
      numericInput("gens", "Generations:", value = 50, min = 1),
      numericInput("sims", "Simulations:", value = 50, min = 1),

      ##Allows the user to select which simulation they want to view
      uiOutput("sim_select"),
      actionButton("run", "Run Simulation", class = "btn-primary")),

    ##Create the main panel for what will be displayed in shiny.
    mainPanel(
      h4("Population Counts: Generation 0, Post-Bottleneck, and Final Generation"),
      tableOutput("final_table"),
      hr(),
      h4("Summary Statistics Across All Simulations (Final Generation)"),
      tableOutput("summary_table"),
      hr(),
      h4("Average Allele Frequency Over Generations and All Simulations"),
      plotOutput("drift_plot"),
      hr(),

      ##Displays the allele frequency for the selected simulation
      h4("Allele Frequency for Selected Simulation"),
      plotOutput("single_sim_plot")
    )
  )
)

server <- function(input, output) {

  ##This runs the entire simulation when the user clicks run
  simulation_data <- eventReactive(input$run, {

    ##User inputs
    population_size <- input$pop_size
    generations <- input$gens
    brown_frequency <- input$brown
    blue_frequency <- input$blue
    green_frequency <- 1 - brown_frequency - blue_frequency

    ##Make sure frequencies are valid
    validate(need(green_frequency >= 0, "Brown + Blue frequencies must sum to 1 or less."))

    ##Create frequency vector
    start_frequencies <- c(brown = brown_frequency, blue = blue_frequency, green = green_frequency)

    ##Create generation 0
    allele_counts <- round(start_frequencies * population_size)
    difference <- population_size - sum(allele_counts)
    if (difference != 0) {
      allele_counts[which.max(allele_counts)] <- allele_counts[which.max(allele_counts)] + difference
    }

    alleles <- rep(names(allele_counts), times = allele_counts)
    alleles <- sample(alleles)

    generation0 <- data.frame(individual = 1:population_size, allele = alleles)

    ##Run bottleneck
    pop <- 1:population_size
    bottle_neck <- remove_random_pct(pop)

    bottleneck_data <- inner_join(generation0, bottle_neck, by = c("individual" = "result")) %>%
      select(-pct_removed)

    ##Calculate new frequencies after bottleneck
    allele_counts_new <- bottleneck_data %>% count(allele)
    new_population_size <- nrow(bottleneck_data)

    new_start_frequencies <- allele_counts_new %>%
      mutate(freq = n / sum(n)) %>%
      select(allele, freq) %>%
      deframe()

    ##Run simulations
    results <- run_simulations(input$sims, new_population_size, new_start_frequencies, generations)

    list(
      results = results,
      new_population_size = new_population_size,
      generation0 = generation0,
      bottleneck_data = bottleneck_data
    )
  })

  ##Creates the dropdown menu for selecting simulations
  output$sim_select <- renderUI({
    req(simulation_data())
    sims <- unique(simulation_data()$results$simulation)

    selectInput("selected_sim", "Select Simulation:",
                choices = sims,
                selected = sims[1])
  })

  ##Create table for generation 0, post-bottleneck, and final generation
  output$final_table <- renderTable({

    generation0 <- simulation_data()$generation0
    bottleneck_data <- simulation_data()$bottleneck_data
    results <- simulation_data()$results
    new_population_size <- simulation_data()$new_population_size

    gen0_counts <- generation0%>%
      count(allele)%>%
      mutate(Stage = "Generation 0", Count = n)%>%
      select(Stage, allele, Count)

    bottleneck_counts <- bottleneck_data%>%
      count(allele) %>%
      mutate(Stage = "Post-Bottleneck", Count= n)%>%
      select(Stage, allele, Count)

    ##Use selected simulation for final generation
    final_counts <- results%>%
      filter(simulation == input$selected_sim)%>%
      filter(generation== max(generation)) %>%
      mutate(Count = round(frequency * new_population_size))%>%
      mutate(Stage = "Final Generation")%>%
      select(Stage, allele, Count)

    bind_rows(gen0_counts, bottleneck_counts, final_counts)
  })

  ##Create summary statistics table
  output$summary_table <- renderTable({
    results <- simulation_data()$results
    summary_stats(results)
  })

  ##Plot average allele frequency across all simulations
  output$drift_plot <- renderPlot({
    results <- simulation_data()$results

    ggplot(results, aes(x = generation, y = frequency, color = allele))+
      stat_summary(fun = mean, geom = "line", linewidth = 1)+
      scale_color_manual(values = c(brown = "brown", blue = "blue", green = "darkgreen"))+
      labs(title = "Average Allele Frequency Over Generations", x = "Generation", y = "Frequency",color = "Allele")+
      theme_minimal()
  })

  ##Plot allele frequency for the selected simulation
  output$single_sim_plot <- renderPlot({
    req(simulation_data(), input$selected_sim)

    results <- simulation_data()$results%>%
      filter(simulation == input$selected_sim)

    ggplot(results, aes(x=generation, y=frequency, color = allele))+
      geom_line(linewidth = 1)+
      scale_color_manual(values=c(brown ="brown", blue = "blue", green = "darkgreen"))+
      labs(title = paste("Simulation", input$selected_sim), x = "Generation", y ="Frequency", color = "Allele")+
      theme_minimal()
  })
}
shinyApp(ui, server)
```

